---
title: "BOLDistilled Library Metadata"
output:
  pdf_document:
    latex_engine: pdflatex
params:
  wd: ""
  report_date: ""
  filename: ""
  threshold: ""
  count: ""
  bin_count: ""
  bin_inel_count: ""
  redundant_otu_count: ""
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{times}  # Set to Times font
  - \usepackage{titlesec}
  - \titlespacing*{\section}{0pt}{5ex}{1ex}  # Adjust space around section
  - \titlespacing*{\subsection}{0pt}{5ex}{1ex}  # Adjust space around subsection
  - \titleformat{\section}[block]{\normalfont\bfseries\Large}{\thesection}{1em}{}
  - \titleformat{\subsection}[block]{\normalfont\bfseries\large}{\thesubsection}{1em}{}
  - \setlength{\parskip}{0pt}  # Remove extra space between paragraphs
  - \setlength{\headheight}{0pt}  # Remove header space
  - \setlength{\topskip}{0pt}  # Remove space before title
  - \setlength{\textheight}{730pt}  # Adjust text height to make use of more page space
  - \usepackage[font={large,rm}, labelfont={bf}, textfont={it}, justification=raggedright, singlelinecheck=false]{caption}
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# v1.1

# extract parameters
wd <- params$wd
report_date <- params$report_date
filename <- params$filename
threshold <- params$threshold
count <- as.numeric(params$count)
bin_count <- as.numeric(params$bin_count)
bin_inel_count <- as.numeric(params$bin_inel_count)
redundant_otu_count <- suppressWarnings(as.numeric(params$redundant_otu_count))
if (length(redundant_otu_count) == 0 || is.na(redundant_otu_count)) redundant_otu_count <- 0

# set working directory
setwd(wd)
library(data.table)
library(ggplot2)
library(dplyr)
library(Biostrings)

input.fasta <- readDNAStringSet(paste0(filename, "_SEQUENCES.fasta"))
num.records <- length(input.fasta)
compression <- 100-round((num.records / count) * 100, 1)
num.records <- prettyNum(num.records, big.mark=",", preserve.width="none")
count <- prettyNum(count, big.mark=",", preserve.width="none")
bin_count <- prettyNum(bin_count, big.mark=",", preserve.width="none")
bin_inel_count <- prettyNum(bin_inel_count, big.mark=",", preserve.width="none")
redundant_otu_count <- prettyNum(redundant_otu_count, big.mark=",", preserve.width="none")

# calculate stats
record.list <- sapply(strsplit(names(input.fasta), "\\|"), "[", 2)

num.records.bins <- length(record.list[grepl("BOLD:", record.list)])
num.records.bins <- prettyNum(num.records.bins, big.mark=",", preserve.width="none")

num.unique.bins <- length(unique(record.list[grepl("BOLD:", record.list)]))
num.unique.bins <- prettyNum(num.unique.bins, big.mark=",", preserve.width="none")

private_bin_df <- data.frame("temp" = names(input.fasta))
private_bin_df$PID <- sapply(strsplit(private_bin_df$temp, "\\|"), "[", 1)
private_bin_df$BIN <- sapply(strsplit(private_bin_df$temp, "\\|"), "[", 2)
private_bin_df <- private_bin_df[,-1]
private_bin_df <- private_bin_df[grepl("BOLD:", private_bin_df$BIN),]
private_bin_df$PUBLIC <- grepl("private", private_bin_df$PID)
private_bin_df$PUBLIC <- ifelse(private_bin_df$PUBLIC, 0, 1)
private_bin_df <- private_bin_df %>%
  group_by(BIN) %>%
  summarise(public_count = sum(PUBLIC)) %>%
  ungroup()

private_only <- private_bin_df[private_bin_df$public_count == 0,]
num.private.bins <- length(unique(private_only$BIN))
num.private.bins <- prettyNum(num.private.bins, big.mark=",", preserve.width="none")

public <- private_bin_df[private_bin_df$public_count != 0,]
num.public.bins <- length(unique(public$BIN))
num.public.bins <- prettyNum(num.public.bins, big.mark=",", preserve.width="none")

num.records.otus <- length(record.list[grepl("OTU:", record.list)])
num.records.otus <- prettyNum(num.records.otus, big.mark=",", preserve.width="none")

num.unique.otus <- length(unique(record.list[grepl("OTU:", record.list)]))
num.unique.otus <- prettyNum(num.unique.otus, big.mark=",", preserve.width="none")

table.bins <- data.frame(table(record.list[grepl("BOLD:", record.list)]))
mean.records.per.bin <- round(mean(table.bins$Freq), digits = 1)

table.otus <- data.frame(table(record.list[grepl("OTU:", record.list)]))
mean.records.per.otu <- round(mean(table.otus$Freq), digits = 1)

num.private.records <- length(input.fasta[grepl("private_record", names(input.fasta))])
num.private.records <- prettyNum(num.private.records, big.mark=",", preserve.width="none")

num.public.records <- length(input.fasta[!grepl("private_record", names(input.fasta))])
num.public.records <- prettyNum(num.public.records, big.mark=",", preserve.width="none")

input.tax <- fread(paste0(filename, "_TAXONOMY.tsv"), header = T, sep = "\t", fill = TRUE)
temp_tax <- data.frame("BIN" = record.list)
temp_tax <- merge(temp_tax, input.tax, by.x = "BIN", by.y = "bin", all.x = TRUE)
temp_tax[temp_tax == ""] <- NA

num.id.to.kingdom <- sum(!is.na(temp_tax$kingdom))
num.id.to.kingdom <- prettyNum(num.id.to.kingdom, big.mark=",", preserve.width="none")

num.id.to.phylum <- sum(!is.na(temp_tax$phylum))
num.id.to.phylum <- prettyNum(num.id.to.phylum, big.mark=",", preserve.width="none")

num.id.to.class <- sum(!is.na(temp_tax$class))
num.id.to.class <- prettyNum(num.id.to.class, big.mark=",", preserve.width="none")

num.id.to.order <- sum(!is.na(temp_tax$order))
num.id.to.order <- prettyNum(num.id.to.order, big.mark=",", preserve.width="none")

num.id.to.family <- sum(!is.na(temp_tax$family))
num.id.to.family <- prettyNum(num.id.to.family, big.mark=",", preserve.width="none")

num.id.to.genus <- sum(!is.na(temp_tax$genus))
num.id.to.genus <- prettyNum(num.id.to.genus, big.mark=",", preserve.width="none")

num.id.to.species <- sum(!is.na(temp_tax$species))
num.id.to.species <- prettyNum(num.id.to.species, big.mark=",", preserve.width="none")

#######################################
##### plot composition of kingdoms ####
#######################################
# calculate total records across all taxa
total_all_records <- nrow(temp_tax)

# summarize counts per kingdom
df_summary <- temp_tax %>%
  group_by(kingdom) %>%
  summarise(total_records = n(), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# replace NA kingdoms with "unresolved"
df_summary <- df_summary %>%
  mutate(kingdom = ifelse(is.na(kingdom), "unresolved", kingdom))

plot.k <- ggplot(df_summary, aes(x = reorder(kingdom, -total_records), y = total_records, fill = kingdom)) + 
  geom_bar(stat = "identity") +  
  scale_y_continuous(
    trans = "log", 
    limits = c(1, max(df_summary$total_records, na.rm = TRUE)), 
    breaks = scales::breaks_log(n = 7), 
    labels = scales::label_comma(),  
    expand = expansion(mult = c(0, 0.2)) 
  ) +  
  geom_text(aes(label = paste0(scales::comma(total_records), "\n(", round(percentage, 1), "%)")), 
            vjust = -0.5, size = 2.5) +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = NULL, y = "Log Count of Records") +
  theme(legend.position = "none",
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        panel.grid.minor = element_blank())


#######################################
##### plot composition of phyla #######
#######################################
# summarize counts per phylum
df_summary_phylum <- temp_tax %>%
  group_by(phylum) %>%
  summarise(total_records = n(), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# identify the top 6 phyla based on total records
top_phyla <- df_summary_phylum %>%
  arrange(desc(total_records)) %>%
  slice_head(n = 6) %>%
  pull(phylum)

# replace less frequent phyla with "Other"
df_summary_phylum <- df_summary_phylum %>%
  mutate(phylum = ifelse(phylum %in% top_phyla, phylum, "Other")) %>%
  group_by(phylum) %>%
  summarise(total_records = sum(total_records), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# ensure "Other" is placed last
df_summary_phylum <- df_summary_phylum %>%
  arrange(desc(percentage)) 

# set factor levels to ensure "Other" is last
df_summary_phylum$phylum <- factor(df_summary_phylum$phylum, 
                                   levels = c(df_summary_phylum$phylum[df_summary_phylum$phylum != "Other"], "Other"))

# plot composition of top 6 phyla + "Other"
plot.p <- ggplot(df_summary_phylum, aes(x = phylum, y = total_records, fill = phylum)) + 
  geom_bar(stat = "identity") +  
  scale_y_continuous(
    trans = "log",  
    limits = c(1, max(df_summary_phylum$total_records, na.rm = TRUE)), 
    breaks = scales::breaks_log(n = 7),  
    labels = scales::label_comma(),  
    expand = expansion(mult = c(0, 0.2))  
  ) +  
  geom_text(aes(label = paste0(scales::comma(total_records), "\n(", round(percentage, 1), "%)")), 
            vjust = -0.5, size = 3) +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = NULL, y = "Log Count of Records") +
  theme(legend.position = "none",
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        panel.grid.minor = element_blank())


##########################################################
##### plot composition of orders within Arthropoda #######
##########################################################
# summarize counts per order within Arthropoda
df_summary_order <- temp_tax %>%
  filter(phylum == "Arthropoda") %>%
  group_by(order) %>%
  summarise(total_records = n(), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# identify the top 6 orders based on total records
top_orders <- df_summary_order %>%
  arrange(desc(total_records)) %>%
  slice_head(n = 6) %>%
  pull(order)

# replace less frequent orders with "Other"
df_summary_order <- df_summary_order %>%
  mutate(order = ifelse(order %in% top_orders, order, "Other")) %>%
  group_by(order) %>%
  summarise(total_records = sum(total_records), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# ensure "Other" is placed last
df_summary_order <- df_summary_order %>%
  arrange(desc(percentage))  

# set factor levels to ensure "Other" is last
df_summary_order$order <- factor(df_summary_order$order, 
                                 levels = c(df_summary_order$order[df_summary_order$order != "Other"], "Other"))

# plot composition of top 6 orders + "Other"
plot.oa <- ggplot(df_summary_order, aes(x = order, y = total_records, fill = order)) + 
  geom_bar(stat = "identity") +  
  scale_y_continuous(
    trans = "log",  
    limits = c(1, max(df_summary_order$total_records, na.rm = TRUE)),  
    breaks = scales::breaks_log(n = 7),  
    labels = scales::label_comma(),  
    expand = expansion(mult = c(0, 0.2))  
  ) +  
  geom_text(aes(label = paste0(scales::comma(total_records), "\n(", round(percentage, 1), "%)")), 
            vjust = -0.5, size = 3) + 
  theme_minimal() +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = NULL, y = "Log Count of Records") +
  theme(legend.position = "none",
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        panel.grid.minor = element_blank())

##########################################################
##### plot composition of orders within Chordata #########
##########################################################
# summarize counts per order within Chordata
df_summary_order_chordata <- temp_tax %>%
  filter(phylum == "Chordata") %>%
  group_by(order) %>%
  summarise(total_records = n(), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100) 

# identify the top 6 orders based on total records
top_orders_chordata <- df_summary_order_chordata %>%
  arrange(desc(total_records)) %>%
  slice_head(n = 6) %>%
  pull(order)

# replace less frequent orders with "Other"
df_summary_order_chordata <- df_summary_order_chordata %>%
  mutate(order = ifelse(order %in% top_orders_chordata, order, "Other")) %>%
  group_by(order) %>%
  summarise(total_records = sum(total_records), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# ensure "Other" is placed last
df_summary_order_chordata <- df_summary_order_chordata %>%
  arrange(desc(percentage))  

# set factor levels to ensure "Other" is last
df_summary_order_chordata$order <- factor(df_summary_order_chordata$order, 
                                          levels = c(df_summary_order_chordata$order[df_summary_order_chordata$order != "Other"], "Other"))

# plot composition of top 6 orders + "Other" within Chordata
plot.oc <- ggplot(df_summary_order_chordata, aes(x = order, y = total_records, fill = order)) + 
  geom_bar(stat = "identity") +  
  scale_y_continuous(
    trans = "log",  
    limits = c(1, max(df_summary_order_chordata$total_records, na.rm = TRUE)),  
    breaks = scales::breaks_log(n = 7),  
    labels = scales::label_comma(),  
    expand = expansion(mult = c(0, 0.2))  
  ) +  
  geom_text(aes(label = paste0(scales::comma(total_records), "\n(", round(percentage, 1), "%)")), 
            vjust = -0.5, size = 3) +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = NULL, y = "Log Count of Records") +
  theme(legend.position = "none",
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        panel.grid.minor = element_blank())

######################################################################
##### plot composition of common vert groups within Chordata #########
######################################################################
# filter for Chordata
chordata_tax <- temp_tax %>%
  filter(phylum == "Chordata")

# define broad categories based on class
chordata_tax <- chordata_tax %>%
  mutate(category = case_when(
    class %in% c("Teleostei", "Elasmobranchii", "Dipneusti") ~ "Fish",
    class == "Aves" ~ "Birds",
    class == "Mammalia" ~ "Mammals",
    class %in% c("Reptilia", "Amphibia") ~ "Reptiles/Amphibians",
    TRUE ~ "Other"
  ))

# summarize counts per category
df_summary_category <- chordata_tax %>%
  group_by(category) %>%
  summarise(total_records = n(), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# identify the top 6 categories based on total records
top_categories <- df_summary_category %>%
  arrange(desc(total_records)) %>%
  slice_head(n = 6) %>%
  pull(category)

# replace less frequent categories with "Other"
df_summary_category <- df_summary_category %>%
  mutate(category = ifelse(category %in% top_categories, category, "Other")) %>%
  group_by(category) %>%
  summarise(total_records = sum(total_records), .groups = "drop") %>%
  mutate(percentage = total_records / total_all_records * 100)  

# ensure "Other" is placed last
df_summary_category <- df_summary_category %>%
  arrange(desc(percentage)) 

# set factor levels to ensure "Other" is last
df_summary_category$category <- factor(df_summary_category$category, 
                                       levels = c(df_summary_category$category[df_summary_category$category != "Other"], "Other"))

# plot composition of top 6 categories + "Other" within Chordata
plot.verts <- ggplot(df_summary_category, aes(x = category, y = total_records, fill = category)) + 
  geom_bar(stat = "identity") +  
  scale_y_continuous(
    trans = "log",  
    limits = c(1, max(df_summary_category$total_records, na.rm = TRUE)),  
    breaks = scales::breaks_log(n = 7),  
    labels = scales::label_comma(),  
    expand = expansion(mult = c(0, 0.2))  
  ) +  
  geom_text(aes(label = paste0(scales::comma(total_records), "\n(", round(percentage, 1), "%)")), 
            vjust = -0.5, size = 3) +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = NULL, y = "Log Count of Records") +
  theme(legend.position = "none",
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        panel.grid.minor = element_blank())
```
\vspace*{-\baselineskip}
# Version Details:
- **Name**: `r filename`
- **Date**: `r report_date`
- **Divergence Threshold**: `r threshold`

# Summary of records:

- **Count of records before BOLDistillation**: `r count`
  - **Assigned to a BIN**: `r bin_count`
  - **BIN-ineligible:**: `r bin_inel_count`
  
- **Count of records after BOLDistillation**: `r paste0(num.records, " (", compression, "% reduction)")`
  - **Number of records assigned to a BIN**: `r num.records.bins`
    - **Number of unique BINs**: `r num.unique.bins`
      - **Number of private BINs**: `r num.private.bins`
      - **Number of public BINs**: `r num.public.bins`
    - **Mean records per BIN**: `r mean.records.per.bin`
  - **Number of records assigned to an OTU**: `r num.records.otus`
    - **Number of unique OTUs**: `r num.unique.otus`
    - **Mean records per OTU**: `r mean.records.per.otu`
    - **OTU reps removed as redundant with a BIN**: `r redundant_otu_count`
  - **Number of public records**: `r num.public.records`
  - **Number of anonymized private records**: `r num.private.records`
  
# Summary of taxonomic composition:

- **Number of records identified to kingdom**: `r num.id.to.kingdom`
- **Number of records identified to phylum**: `r num.id.to.phylum`
- **Number of records identified to class**: `r num.id.to.class`
- **Number of records identified to order**: `r num.id.to.order`
- **Number of records identified to family**: `r num.id.to.family`
- **Number of records identified to genus**: `r num.id.to.genus`
- **Number of records identified to species**: `r num.id.to.species`

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Number of BOLDistilled records belonging to each kingdom."}
print(plot.k)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Number of BOLDistilled records belonging to each phylum."}
print(plot.p)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Number of BOLDistilled records belonging to each arthropod order."}
print(plot.oa)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Number of BOLDistilled records belonging to each vertebrate order."}
print(plot.oc)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Number of BOLDistilled records belonging to common vertebrate groups."}
print(plot.verts)
```

